---
title: "Online Shoppers Purchasing Intention: Exploratory Analysis and Logistic Regression"
author: "Riya Dutta"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include = FALSE}
# Show code in the report, but hide package messages and warnings
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Run all code from the project folder (one level up from report/), so that
# file paths such as "data/raw/..." work
knitr::opts_knit$set(root.dir = "..")
```

## 1. Introduction

This report examines how visitors browse an online shop and which browsing behaviours are associated with completing a purchase. The data are observational, so all results describe associations and should not be interpreted as causal effects.

## 2. Data

The analysis uses the Online Shoppers Purchasing Intention Dataset (Sakar and Kastro, 2018). Each row is one browsing session, and the outcome variable `Revenue` records whether the session ended in a purchase.

| Variable | Description |
|---|---|
| `Administrative`, `Informational`, `ProductRelated` | Number of pages of each type visited in the session |
| `Administrative_Duration`, `Informational_Duration`, `ProductRelated_Duration` | Total time spent on each page type, in seconds |
| `BounceRates` | Average bounce rate of the pages visited (share of visitors who entered on that page and left without further interaction) |
| `ExitRates` | Average exit rate of the pages visited (share of page views that were the last in a session) |
| `PageValues` | Average Google Analytics page value of the pages visited, which is calculated from pages that preceded completed transactions |
| `SpecialDay` | Closeness of the visit date to a special day such as Valentine's Day, from 0 (not close) to 1 |
| `Month` | Month of the visit |
| `OperatingSystems`, `Browser`, `Region`, `TrafficType` | Category codes for the visitor's technology, location and traffic source |
| `VisitorType` | New, returning or other visitor |
| `Weekend` | Whether the visit took place at the weekend |
| `Revenue` | Whether the session ended in a purchase (outcome) |

The data are prepared by `scripts/02_clean.R`, which sets the correct variable types. The 125 duplicate rows are kept, because they are consistent with separate very short visits rather than recording errors.

```{r load-data, results = "hide"}
# Load the tidyverse and create the cleaned table called shoppers
library(tidyverse)
source("scripts/02_clean.R")
```

## 3. Exploratory analysis

### 3.1 Browsing behaviour by purchase outcome

The table compares the median and mean of each browsing measure between sessions that did and did not end in a purchase. Medians are reported alongside means because most measures are strongly right-skewed: many sessions have values of zero and a few have very large values.

```{r browsing-summary}
browsing_summary <- shoppers |>
  # Label the outcome in words
  mutate(Outcome = if_else(Revenue, "Purchase", "No purchase")) |>
  # Reshape so that each row holds one measure for one session
  pivot_longer(Administrative:SpecialDay, names_to = "Measure", values_to = "value") |>
  # Keep the measures in their original column order
  mutate(Measure = fct_inorder(Measure)) |>
  # Calculate the median and mean of each measure for each outcome
  group_by(Measure, Outcome) |>
  summarise(Median = median(value), Mean = mean(value), .groups = "drop") |>
  # Place the two outcomes side by side
  pivot_wider(names_from = Outcome, values_from = c(Median, Mean), names_sep = ": ")

knitr::kable(browsing_summary, digits = 3)
```

On average, sessions that ended in a purchase involved more pages and more time on every page type, and had lower bounce and exit rates, than sessions that did not. Most sessions in both groups included no informational pages, so the medians for these two measures are zero. The largest difference is in `PageValues`: 89% of sessions without a purchase have a page value of zero, compared with 19% of sessions with a purchase. Because page value is calculated from pages that preceded completed transactions, this variable is partly derived from the outcome, which needs to be considered before it is used in a model.

### 3.2 Purchase rate by visitor characteristics

The purchase rate is the share of sessions in each group that ended in a purchase. Across all sessions it is 15.5%.

```{r rate-visitor}
shoppers |>
  group_by(VisitorType) |>
  summarise(Sessions = n(), Purchases = sum(Revenue), `Purchase rate` = mean(Revenue)) |>
  knitr::kable(digits = 3)
```

```{r rate-weekend}
shoppers |>
  group_by(Weekend) |>
  summarise(Sessions = n(), Purchases = sum(Revenue), `Purchase rate` = mean(Revenue)) |>
  knitr::kable(digits = 3)
```

```{r rate-month}
shoppers |>
  group_by(Month) |>
  summarise(Sessions = n(), Purchases = sum(Revenue), `Purchase rate` = mean(Revenue)) |>
  knitr::kable(digits = 3)
```

New visitors had a higher purchase rate (24.9%) than returning visitors (13.9%), and weekend sessions had a slightly higher rate (17.4%) than weekday sessions (14.9%). Purchase rates varied by month, from 1.6% in February to 25.4% in November. The number of sessions also differs widely between months, and January and April are absent from the data, so monthly patterns should be read with caution.

### 3.3 Visualising the differences

Figure 1 shows the distribution of three browsing measures for each outcome. The vertical axes use a log(1 + x) scale, which spreads out the many small values and compresses the few very large ones, so that both groups can be compared on the same plot. Exit rates are on a much smaller scale and are summarised in the table in Section 3.1.

```{r fig-browsing, fig.width = 8, fig.height = 4}
# Colours used for the two outcomes in all figures
outcome_colours <- c("No purchase" = "#2a78d6", "Purchase" = "#eb6834")

shoppers |>
  mutate(Outcome = if_else(Revenue, "Purchase", "No purchase")) |>
  # Reshape so that the three measures can be drawn side by side
  pivot_longer(c(ProductRelated, ProductRelated_Duration, PageValues),
               names_to = "Measure", values_to = "value") |>
  mutate(Measure = fct_inorder(Measure)) |>
  # One box plot per outcome, coloured by outcome
  ggplot(aes(x = Outcome, y = value, fill = Outcome)) +
  geom_boxplot(width = 0.5, outlier.size = 0.6, outlier.alpha = 0.2) +
  # One panel per measure, each with its own vertical scale
  facet_wrap(~ Measure, nrow = 1, scales = "free_y") +
  # Log(1 + x) scale, with axis marks at 0, 1, 10, 100, 1,000 and 10,000
  scale_y_continuous(transform = "log1p", breaks = c(0, 1, 10, 100, 1000, 10000),
                     labels = scales::label_comma()) +
  scale_fill_manual(values = outcome_colours) +
  labs(title = "Figure 1. Browsing measures by purchase outcome",
       x = NULL, y = "Value (log(1 + x) scale)") +
  theme_minimal() +
  # No legend (the horizontal axis already names each outcome) and no minor gridlines
  theme(legend.position = "none", panel.grid.minor = element_blank())
```

Purchase sessions tend to include more product pages and more time on them, although the two distributions overlap considerably. The contrast is much sharper for `PageValues`: most sessions without a purchase have a page value of zero, whereas most purchase sessions have a positive value, consistent with the way this measure is calculated.

Figure 2 shows the purchase rate in each month. The dashed line marks the rate across all sessions.

```{r fig-month, fig.width = 8, fig.height = 4}
# Purchase rate across all sessions, used for the reference line
overall_rate <- mean(shoppers$Revenue)

shoppers |>
  group_by(Month) |>
  summarise(purchase_rate = mean(Revenue)) |>
  # One bar per month
  ggplot(aes(x = Month, y = purchase_rate)) +
  geom_col(fill = "#2a78d6", width = 0.6) +
  # Reference line and its label
  geom_hline(yintercept = overall_rate, linetype = "dashed", colour = "grey40") +
  annotate("text", x = 0.6, y = overall_rate, vjust = -0.6, hjust = 0, size = 3.5,
           colour = "grey30", label = paste0("All sessions: ", round(overall_rate * 100, 1), "%")) +
  scale_y_continuous(labels = scales::label_percent()) +
  labs(title = "Figure 2. Purchase rate by month", x = NULL, y = "Purchase rate") +
  theme_minimal() +
  # Keep only the horizontal gridlines
  theme(panel.grid.minor = element_blank(), panel.grid.major.x = element_blank())
```

Purchase rates were below the overall rate from February to June and above it from August to November, peaking in November. As noted above, months differ greatly in their number of sessions, so this pattern describes this particular sample and may not reflect seasonal behaviour in general.

## 4. Logistic regression

### 4.1 Main model

A logistic regression models the probability that a session ends in a purchase as a function of several session characteristics at once. Each estimate describes the association between one characteristic and the outcome while the other characteristics in the model are held constant.

The predictors were chosen as follows:

- **Page counts** for administrative, informational and product pages are included. The matching duration measures are left out, because they overlap strongly with the counts (for product pages, the correlation between count and duration is 0.86), and including both would make the estimates unstable.
- **Exit rate** is included and **bounce rate** is left out, because the two are very highly correlated (0.91).
- **Special day, month, visitor type and weekend** are included.
- **Operating system, browser, region and traffic type** are left out. They are anonymised technical codes with many small categories and no clear behavioural interpretation.
- **Page value** is left out of the main model, because it is partly calculated from completed purchases (Section 3.1). Section 4.3 compares a model that includes it.

Two predictors are rescaled so that their estimates are easier to read: product pages are counted in units of 10 pages, and exit rate is expressed in percentage points. The reference groups for the categorical predictors are May and returning visitors, the largest group in each case.

```{r model-main}
# Prepare the data for the model
model_data <- shoppers |>
  mutate(
    # Product pages in units of 10, and exit rate in percentage points
    ProductRelated_10 = ProductRelated / 10,
    ExitRates_pct = ExitRates * 100,
    # Compare each month with May, and each visitor type with returning visitors
    Month = fct_relevel(Month, "May"),
    VisitorType = fct_relevel(VisitorType, "Returning_Visitor")
  )

# Fit the logistic regression (family = binomial means the outcome is TRUE or FALSE)
model_main <- glm(
  Revenue ~ Administrative + Informational + ProductRelated_10 + ExitRates_pct +
    SpecialDay + Month + VisitorType + Weekend,
  family = binomial,
  data = model_data
)

# Turn the results into a table of odds ratios with 95% confidence intervals
library(broom)
tidy(model_main, exponentiate = TRUE, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  transmute(
    Predictor = term,
    `Odds ratio` = estimate,
    `95% CI lower` = conf.low,
    `95% CI upper` = conf.high,
    `p-value` = scales::label_pvalue()(p.value)
  ) |>
  knitr::kable(digits = 2)
```

An odds ratio above 1 means that higher values of the predictor (or membership of the group, compared with the reference group) are associated with higher odds of a purchase, and an odds ratio below 1 means lower odds. The 95% confidence interval shows the range of values compatible with the data; an interval that does not include 1 indicates an association that is unlikely to be due to chance alone.

Holding the other predictors constant:

- **Exit rate** shows the strongest association. Each additional percentage point in the average exit rate of the pages visited is associated with about 26% lower odds of a purchase (odds ratio 0.74).
- **Product pages** and **informational pages** are both associated with slightly higher odds of a purchase. The association for administrative pages is small and its confidence interval includes 1.
- **New visitors** had about 42% higher odds of a purchase than returning visitors.
- Compared with May, sessions in **November** had about twice the odds of a purchase, and sessions in **February** and **March** had lower odds. The February estimate is based on only three purchases, which is reflected in its very wide confidence interval.
- Sessions closer to a **special day** had lower odds of a purchase.
- There was no clear difference between **weekend** and weekday sessions once the other predictors were taken into account.

These are associations in observational data. They do not show that changing any of these characteristics would change the likelihood of a purchase.

## References

Sakar, C. and Kastro, Y. (2018). *Online Shoppers Purchasing Intention Dataset* [Dataset]. UCI Machine Learning Repository. https://doi.org/10.24432/C5F88Q. Licensed under CC BY 4.0.
